Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-07-27 manuscript/manuscript.tex
adding Frank's change requests

\newcommand{\versionnumber}{1.0} % TODO: can we extract this number from a git tag?
\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}

@@

For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail.

In addition to publications on theory and application of Markov state modeling~\cite{msm-book,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.

@@

A spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse-grain the $75$ $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
\[ \begin{array}{ccc}
\textrm{macrostate } \mathcal{S}_i & \pi_{\mathcal{S}_i} & G_{\mathcal{S}_i} / \textrm{k}_\textrm{B} T \\
\hline
\mathcal{S}_1 & 0.004 & 5.567 \\
\mathcal{S}_2 & 0.014 & 4.293 \\
\mathcal{S}_3 & 0.021 & 3.841 \\
\mathcal{S}_4 & 0.021 & 3.875 \\
\mathcal{S}_5 & 0.940 & 0.062 \\
\end{array}\]
using the relation
\begin{equation}
\label{eq:fe}
G_{\mathcal{S}_i} = - \textrm{k}_\textrm{B} T \ln \sum\limits_{j\in \mathcal{S}_i} \pi_j,
\end{equation}
where $\pi_j$ denotes the MSM stationary weight of the $j^\textrm{th}$ microstate.

@@

>>>>

\newcommand{\versionnumber}{0.3} % TODO: can we extract this number from a git tag?
\newcommand{\githubrepository}{\url{github.com/markovmodel/pyemma_tutorials}}

@@

For those unfamiliar with Markov state modeling, ``\emph{Markov State Models: From an Art to a Science}''~\cite{msm-brooke} provides a recent overview, while ``\emph{Markov models of molecular kinetics: Generation and validation}''~\cite{msm-jhp} describes the basic MSM theory and methodology in detail. Two textbooks exist that focus on computational methods and applications~\cite{msm-book} and mathematical theory~\cite{schuette-sarich-book}.

In addition to publications on theory and application of Markov state modeling~\cite{schuette-msm,buchete-msm-2008,noe-tmat-sampling,bowman-msm-2009,noe-folding-pathways,sarich-msm-quality,noe-fingerprints,noe-dy-neut-scatt,Chodera2014,ben-rev-msm,simon-mech-mod-nmr,oom-feliks},
we also recommend the literature on time-lagged independent component analysis (TICA)~\cite{tica,tica3,tica2,kinetic-maps}, transition path theory (TPT)~\cite{weinan-tpt,metzner-msm-tpt},
hidden Markov state models (HMMs)~\cite{noe-proj-hid-msm,hmm-baum-welch-alg,hmm-tutorial}, and variational techniques~\cite{noe-vac,vamp-preprint,gmrq}, as these topics play important roles within the standard MSM workflow.

@@

A spectral clustering using the PCCA++ algorithm~\cite{pcca++,Deuflhard2005-pcca,Kube2007-pcca+} allows us to coarse-grain the $75$ $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
\[ \begin{array}{ccc}
\textrm{macrostate } \mathcal{S}_i & \pi_{\mathcal{S}_i} & G_{\mathcal{S}_i} / \textrm{k}_\textrm{B} T \\
\hline
\mathcal{S}_1 & 0.004 & 5.567 \\
\mathcal{S}_2 & 0.014 & 4.293 \\
\mathcal{S}_3 & 0.021 & 3.841 \\
\mathcal{S}_4 & 0.021 & 3.875 \\
\mathcal{S}_5 & 0.940 & 0.062 \\
\end{array}\]
using the relation
\begin{equation}
\label{eq:fe}
G_{\mathcal{S}_i} = - \textrm{k}_\textrm{B} T \ln \sum\limits_{j\in \mathcal{S}_i} \pi_j,
\end{equation}
where $\pi_j$ denotes the MSM stationary weight of the $j^\textrm{th}$ microstate.

@@

\subsection{Advanced Methods}

While the present tutorial covers Markov State Modeling 101, we encourage the user to explore other, more recent extensions of the methodology. Multi-ensemble Markov models (MEMMs)~\cite{dtram,tram} can be used to combine unbiased and biased simulations so as to probe kinetics of very rare events~\cite{trammbar}; MEMMs are implemented in PyEMMA. Recently, there have been steps towards replacing the traditional user-directed pipeline (involving featurizing, reducing dimension, discretizing, MSM estimation and coarse-graining) by a single end-to-end deep learning method such as VAMPnets~\cite{vampnet}. Other deep learning methods for performing the dimension reduction~\cite{tae}, finding reaction coordinates for enhanced sampling~\cite{hernandez-vde,Sultan2018-vde-enhanced-sampling,Ribeiro2018-rave}, and generative MSMs~\cite{deep-gen-msm-preprint} have been put forward and are likely to spawn an active field of research on its own right. Implementations of some of these methods are available or is under development in the package deeptime \url{github.com/markovmodel/deeptime}.
